Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis

### 2023-03-05 analysis/paper/paper.Rmd
minor edits

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-Chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or species across areas during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new practices. It is widely accepted that mixed farming of rice/millet was introduced into southwest China from central and northern China around 4700 BP. However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records, providing a unique opportunity to understand the local adoption of agriculture. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarker identification and compounds-specific stable isotope analysis. The results show the signals of terrestrial animals for most potsherds, with one showing a C3 plant source during the Neolithic transition. Our case study indicates that hunting might have been the primary source of subsistence, supplemented by rice/millet farming. This study also demonstrates a perspective from organic residue analysis to explore pottery use and food practices during the Neolithic period. A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diets and subsistence.

@@

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000]. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 4,700 BP and to western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022]. As a result, the Neolithic Hengduan Mountain areas are critical for understanding agricultural expansion processes. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the presence of millet and rice at around the same time as mixed farming appeared in Sichuan [@Chen2015; @Hao2022; @Weisskopf2015]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

To explore whether the new crops or species became important food sources during the Neolithic transition in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a]. The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022]. Although the analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period. To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use? Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic transition? We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels.

@@

<!--Yanyuan area and Guijiabao site-->
We focus on Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)). Guijiabao is unique for understanding the early adoption of agriculture due to its rarity, which contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming Dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022]. For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles [@Hao2022].

<!--the major finding at GJB-->
The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins. The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a]. Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan and southern part of the Hengduan Mountains [@Guijiabao2017a]. Plant remains reveal the cultivation of crops dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on flotation samples collected from ash pits and stratigraphic layers that can be traced back to 5000 BP [@Guijiabao2016; @Hao2022]. This corresponds to a C4-based diet based on isotopic analysis of four human bones [@Lin2022]. In addition to millet, a smaller number of rice grains (*Oryza sativa*) have also been found in ash pits, where the earliest one dates to 4700 BP [@Hao2022]. A detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences, suggesting a mixed millet/rice cultivation prior to 5000 BP [@Huan2022].

<!--info about Gaoshan-->
To compare culinary practices for different ecological zones, we examine a sample from the Gaoshan site in the southwestern Chengdu Plain, an alluvial plain in the western Sichuan Basin at the foot of the Qinghai-Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at Gaoshan reveal abundant materials, including ceramics, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to similar pottery forms and styles, Gaoshan can be categorized into the early Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Gaoshan2017]. The primary subsistence strategy at Gaoshan could be rice cultivation with pig husbandry, supplemented by millet cultivation [@JiangGaoshan2017; @Lee2020; @He2020]. It is suggested that the spread of rice farming into the Sichuan area occurred at least 4700 years ago from the Middle and Lower Yangzi valleys [@Zhang2010].

@@

## Samples
We selected ceramic fragments that were identified as typical types from Guijiabao and Gaoshan. The fragments were from known contexts with radiocarbon dates: Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)). Ceramic fragments from Guijiabao were recovered from trench 13 excavated during the 2017 field season, along with one fragment from the 2016 season. The pottery samples are mostly gray base fragments with fine sand temper, of which a few have impressed decorations. Although their original forms could not be recovered, the evidence of chars on some fragments and their contexts in cultural layers suggest that those ceramics were used in daily life [@Hao2022, @Guijiabao2017a]. This research also attempts to confirm that those ceramics were used as cooking/storage wares for food.

@@

To identify the sources of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions δ^13^C values of C16:0 and C18:0 using GC-C-IRMS. The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008]. To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C(δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013]. The carbon isotope values for C16:0 and C18:0 distribute between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰. The Δ13C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].

@@

# Discussion
<!--A brief summary of interesting results-->
The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets), C3 plants (e.g., rice), or new species (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period. Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potsherd that might originate from C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that those ceramics were used as cooking or storage pots. The low yield could be attributed to the base parts that we selected for analysis, which may absorb fewer residues than the rim of a vessel [@Evershed2008experimental; @Kwak2020; @Charters1993]. But we also found a base sample dating to the early Neolithic period contained a rich composition of long-chain fatty acids, reflecting good preservation. We recognize limitations and challenges for interpreting lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils. Also, a detailed analysis for faunal remains from Guijiabao is still in progress, which constrains the comparison between different lines of evidence. However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices in Guijiabao.

<!--discussion about possible plant origins-->
While it has been indicated that millets (C4 plants), including foxtail and broomcorn millets, were dominated cereal according to recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022], we did not detect any lipid residues associated with C4 plants in our samples. Surprisingly, we identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020]. It is also possible that native wild weeds classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet, but those seeds only account for a small part of the plant remain according to flotation results [@Guijiabao2016]. Despite stone knives and tools were also found at the sites, it is uncertain if they were used for the practice of agricultural harvesting [@Zhou2019; @Hao2022; @Lin2022]. Another possible source for non-ruminant fats could be freshwater fish, which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b]. However, we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid, from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].

<!--discussion about possible ruminant fat origin-->
Ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from Guijiabao and Gaoshan could be derived from deer or horses according to unearthed faunal remains from the sites, which were accompanied with stone tools for hunting, such as stone balls and arrowheads [@Hao2022; @Guijiabao2020a]. However, whether horse was a part of local food choices during the Neolithic period is uncertain. First, the Pleistocene and Neolithic sites with wild horse bones are primarily concentrated in northern China with only sporadic presence in the south [@Yuan2006]. Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period we focus on. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with the values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002]. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were food sources during this period [@Zhang2016].

We observed a mixture of ruminant fats and C3 plants based on a number of Guijiabao potsherds dating to the historical period (ca. 1000 BP). It appears that C3 plants could be processed more in the historical period compared to the early Neolithic. The source of C3 plants might also include wheat since it was introduced in the historical period [@JiangGaoshan2017]. More studies and larger sample size are needed to explore whether there is a shift in diet over time. The overall depleted δ^13^C values for most potsherds suggest that C3 plants might be an important food source for local animals and could also be included in the local diet at Guijiabao during the Neolithic period. Despite no C4 plant signal being found, one Guijiabao potsherd demonstrates the source of C4 plant-fed ruminants. We also note that that our finding of C3 plant signal based on ceramic residues is not consistent with the result of a C4-based diet according to isotopic analysis of human bones [@Lin2022]. The contradiction could be caused by different sampling strategies yielding contradictory results or by residue analysis reflecting only a portion of the content of food practices. Nevertheless, all pieces of evidence support the existing theory of mixed rice/millet farming in southwest China. Our results indicate that farming practices may not have been adopted as much as we initially thought.

In general, during the Neolithic period, the food source detected from our pottery samples consisted mainly of ruminant fats, which are most likely from local wild animals such as deer. Unearthed stone tools, including arrowheads and balls, also suggest hunting activities. Considering the abundance of wild animal resources in the natural environment of the Hengduan mountain ranges suggested by the case of Yingpanshan site [@He2020], a subsistence relying on hunting to a certain degree could be expected. Crop cultivation could be adopted as a supplement to local food sources while it was introduced in the Neolithic period, and it might only account for a small proportion. The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, which could provide more details about local animal species for comparison. An organic residue approach may only provide a narrow perspective on understanding local diets, but it could improve our current understanding of food sources through direct evidence cooked in ceramics. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in the local diet during the Neolithic period.

# Conclusion
Our case study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids but in low concentrations for most potsherds. The food cooked in those ceramics included ruminant fats and terrestrial plants during the Neolithic transition, when the mixed farming of rice/millet spread into southwest China. It is most likely the terrestrial plant signal from C3 plant, implying rice processing in the pot. Surprisingly, the direct processing of either C3 or C4 plants is not as much as we expected, according to the relatively large quantity of plant remains found at Guijiabao, such as millet grains. Instead, ruminant fats from local wild animals, likely deer, accounted for a major part of early Neolithic food processing. This indicates that the proportion of rice/millet in the local diet was not high during the Neolithic transition, suggesting that hunting might have been an important part of local subsistence, supplemented by rice/millet farming. We recognize the limitations of our interpretation of the local diet due to the small sample size. Nevertheless, our study could offer a perspective from organic residue analysis to explore food preferences and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period. A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.

# Acknowledgements
We thank the Chengdu Municipal Institute of Cultural Relics and Archaeology for their permission to access and analyze the pottery samples. We thank Julian Sachs in the Department of Oceanography, University of Washington for his support in letting us use their laboratory for sample preparation and conducting lipid extraction and analysis. We thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines. We also want to express our gratitude to the anonymous reviewers for their comments that help us improve this paper.

>>>>

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or species across areas during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new practices. It is widely accepted that mixed farming of rice/millet was introduced into southwest China from central and northern China around 4700 BP. However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records, providing a unique opportunity to understand the local adoption of agriculture. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarker identification and compounds-specific stable isotope analysis. The results show the signals of terrestrial animals for most potsherds, with one showing a C3 plant source during the Neolithic transition. Our case study indicates that hunting might have been the primary source of subsistence, supplemented by rice/millet farming. This study also demonstrates a perspective from organic residue analysis to explore pottery use and food practices during the Neolithic period. A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diets and subsistence.

@@

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000]. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 4,700 BP and to western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022]. As a result, the Neolithic Hengduan Mountain areas are critical for understanding agricultural expansion processes. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the presence of millet and rice at around the same time as mixed farming appeared in Sichuan [@Chen2015; @Hao2022; @Weisskopf2015]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

To explore whether the new crops or species became important food sources during the Neolithic transition in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a]. The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought, and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022]. Although the analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period. To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use? Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic transition? We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels.

@@

<!--Yanyuan area and Guijiabao site-->
We focus on Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)). Guijiabao is unique for understanding the early adoption of agriculture due to its rarity, which contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022]. For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles [@Hao2022].

<!--the major finding at GJB-->
The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins. The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a]. Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan and the southern part of the Hengduan Mountains [@Guijiabao2017a]. Plant remains reveal the cultivation of crops dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on flotation samples collected from ash pits and stratigraphic layers that can be traced back to 5000 BP [@Guijiabao2016; @Hao2022]. This corresponds to a C4-based diet based on isotopic analysis of four human bones [@Lin2022]. In addition to millet, a smaller number of rice grains (*Oryza sativa*) have also been found in ash pits, where the earliest one dates to 4700 BP [@Hao2022]. A detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences, suggesting a mixed millet/rice cultivation prior to 5000 BP [@Huan2022].

<!--info about Gaoshan-->
To compare culinary practices for different ecological zones, we examined a sample from the Gaoshan site in the southwestern Chengdu Plain, an alluvial plain in the western Sichuan Basin at the foot of the Qinghai-Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at Gaoshan reveal abundant materials, including ceramics, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to similar pottery forms and styles, Gaoshan can be categorized into the early Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Gaoshan2017]. The primary subsistence strategy at Gaoshan could be rice cultivation with pig husbandry, supplemented by millet cultivation [@JiangGaoshan2017; @Lee2020; @He2020]. It is suggested that the spread of rice farming into the Sichuan area occurred at least 4700 years ago from the Middle and Lower Yangzi valleys [@Zhang2010].

@@

## Samples
We selected ceramic fragments that were identified as typical types from Guijiabao and Gaoshan. The fragments were from known contexts with radiocarbon dates: Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)). Ceramic fragments from Guijiabao were recovered from trench 13 excavated during the 2017 field season, along with one fragment from the 2016 season. The pottery samples are mostly gray base fragments with fine sand temper, of which a few have impressed decorations. Although their original forms could not be recovered, the evidence of chars on some fragments and their contexts in cultural layers suggest that those ceramics were used in daily life [@Guijiabao2017a; @Hao2022]. This research also attempts to confirm that those ceramics were used as cooking/storage wares for food.

@@

To identify the sources of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions, δ^13^C values of C16:0 and C18:0, using GC-C-IRMS. The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008]. To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C(δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013]. The carbon isotope values for C16:0 and C18:0 distribute between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰. The Δ13C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].

@@

# Discussion
<!--A brief summary of interesting results-->
The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets), C3 plants (e.g., rice), or new species (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period. Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify non-ruminant fats in a Guijiabao potsherd that might originate from C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that those ceramics were used as cooking or storage pots. The low yield could be attributed to the base parts that we selected for analysis, which may absorb fewer residues than the rim of a vessel [@Evershed2008experimental; @Kwak2020; @Charters1993]. But we also found a base sample dating to the Neolithic period contained a rich composition of long-chain fatty acids, reflecting good preservation. We recognize limitations and challenges for interpreting lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils. Also, a detailed analysis of faunal remains from Guijiabao is still in progress, which constrains the comparison between different types of evidence. However, the general pattern of lipid profiles could still provide informative clues for a better understanding of culinary practices in Guijiabao.

<!--discussion about possible plant origins-->
While it has been indicated that millets (C4 plants), including foxtail and broomcorn millets, were the dominant cereals according to recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022], we did not detect any lipid residues associated with C4 plants in our samples. Surprisingly, we identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plants/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020]. Based on the flotation results from Guijiabao, it is possible that the source of C3 plants is rice or native wild weeds, such as pennisetum and chenopodiaceae; however, the seeds of wild weeds only account for a small part of the plant [@Guijiabao2016]. Despite stone knives and tools were also found at the sites that might relate to farming practices, more studies are needed to confirm their usage for harvesting crops [@Zhou2019; @Hao2022; @Lin2022]. Another possible source for non-ruminant fats could be freshwater fish, which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b]. But we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid, from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].

<!--discussion about possible ruminant fat origin-->
Ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from Guijiabao and Gaoshan could be derived from deer or horses, according to unearthed faunal remains from the sites and the local natural environments. Stone artifacts, including stone balls and arrowheads, were also found at the site, indicative of hunting activities [@Hao2022; @Guijiabao2020a]. However, whether horse was part of local food choices in southwest China around 4500 years ago is skeptical. First, the Pleistocene and Neolithic sites with wild horse bones are primarily concentrated in northern China, with only a sporadic presence in the south [@Yuan2006]. Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period we focus on. In addition, our sample (SYG-TN13-E22-2#1) shows an isotopic value consistent with the values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002]. Thus, deer is the most likely part of the local food choices. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were food sources during this period [@Zhang2016].

We observed a mixture of ruminant fats and C3 plants based on a number of Guijiabao potsherds dating to the historical period (ca. 1000 BP). It appears that C3 plants could be processed more in the historical period compared to the Neolithic period. In addition to rice, the source of C3 plants could also include wheat since it was introduced during the historical period [@JiangGaoshan2017]. A larger sample size are needed to explore whether there is a shift in diet over time. The overall depleted δ^13^C values for potsherd samples associated with ruminant fats suggest that C3 plants might be an important food source for local animals. Despite direct C4 plant signals not being found, one Guijiabao potsherd demonstrates the source of C4 plant-fed ruminants during the Neolithic period. This indicates either C3 or C4 plants could be the food source for ruminants, but whether they were domesticated is unknown. The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, which could provide more details about local animal species for comparison.

In general, during the Neolithic period, the food source detected in our pottery samples consisted mainly of ruminant fats, which are most likely from local wild animals such as deer. Considering the abundance of wild animal resources in the natural environment of the Hengduan mountain ranges, as suggested by the case of the Yingpanshan site [@He2020], a subsistence relying on hunting to a certain degree could be expected. Crop cultivation could be adopted as a supplement to local food sources while it was introduced in the Neolithic period, but it might only account for a small proportion. For the dietary pattern at Guijiabao previous study suggested a C4-based diet according to the isotopic analysis of human bones [@Lin2022], but we only identified a C3 plant signal from one pottery sample without any signal directly from C4 plants. This inconsistency could result from different sampling strategies or from the sample size issue, where our results only reflect a small portion of food content. Nevertheless, our results suggest that farming practices during the Neolithic period may not have been adopted as much as we previously thought. An organic residue approach may only provide a narrow perspective on local diets, but it could improve our current understanding of food practices through direct evidence that has been cooked in ceramics. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region would offer insight into temporal and regional variations in the local diet during the Neolithic period.

# Conclusion
Our case study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids that could be associated with food types. The food processed in the ceramics included ruminant fats and terrestrial plants during the Neolithic period, when millet and rice spread into southwest China. The source of terrestrial plants could be C3 plants, implying the processing of rice or other wild weeds in the pot. However, the evidence of direct processing of either C3 or C4 plants is not in proportion to the large quantity of plant remains found at Guijiabao, especially millet grains. Instead, ruminant fats from local animals, such as wild deer, could have been a common food source during the early Neolithic. This suggests that hunting might have been an important part of local subsistence, supplemented by rice/millet farming when they were introduced. Our results from Guijiabao might represent a subsistence strategy for adapting to the natural environment of the Hengduan Mountains. We recognize the limitations of our interpretation of the local diet due to the small sample size. Nevertheless, our study could offer a perspective from organic residue analysis to explore food preferences and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period. A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.

# Acknowledgements
We thank the Chengdu Municipal Institute of Cultural Relics and Archaeology for their permission to access and analyze the pottery samples. We thank Julian Sachs in the Department of Oceanography at the University of Washington for his support in letting us use their laboratory for sample preparation and conducting lipid extraction and analysis. We thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines. We also want to express our gratitude to the anonymous reviewers for their comments that helped us improve this paper. This research is sponsored by the National Science and Technology Council (PE11006-0107).
